The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-28 rational.tex
a bit of random stuff for the intro

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{mdwlist}
\usepackage{txfonts}
\usepackage{xspace}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}
\usepackage{bm}
\usepackage{cmll}

@@

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^*
%! ^^^ = ^{-1}
%! eps* = \epsilon^*
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

In 1989, Filinski~\cite{Filinski:1989:DCI:648332.755574} observed that values
and continuations are dual notions. This observation was followed by numerous
papers~\cite{Griffin:1989:FNC:96709.96714, curien00duality, Wadler:2003,
  DBLP:conf/rta/Wadler05} which explored and refined this duality. Even
earlier, Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2}
introduced a logic which contains a dual to implication. Her work has been
distilled in the form of \emph{subtractive logic}~\cite{Crolard01} which has
recently been related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.

@@

We show a deep symmetry between functions and delimited continuations,
values and continuations that arises in {{langRev}} in a manner that
is reminiscent of Filinksi's Symmetric \lcal
~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
extending {{langRev}} with a notion of additive duality over the
monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
Compact Closed Categories. The resulting dual types, which we denote
{{-b}}, have a time traveling ``backward information flow''
interpretation and allow for the encoding of higher-order function and
iteration via the construction of {{trace}} operators, thereby making
the extended language {{langRevEE}} a Turing-complete reversible
programming language with higher-order functions and first-class
delimited continuations.

@@

The goal is to produce a value of type $(\texttt{bool} \times
\texttt{int})$. One part of the program, $e_1$, knows how to produce the
value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
to produce the value of type $\texttt{bool}$ (say {{true}}). Computationally,
we model this as follows. The first subcomputation $e_1$ produces {{((),3)}}
and uses the isomorphism to convert {{()}} to {{(alpha,1/alpha)}} where
{{alpha}} is a yet-unknown boolean value. Reshuffling the components, $e_1$
produces {{((alpha,3), 1/alpha)}}. Similarly, $e_2$ can independently produce
{{((true,beta),1/beta)}} where {{beta}} is a yet-unknown integer value.
When the two values meet, we can group the components as follows:
\[
{{(alpha,beta)}}  {{(3,1/beta)}}  {{(true,1/alpha)}}
\]
Using the isomorphism in the right-to-left direction on the last two pairs,
forces {{alpha}} to be resolved to {{true}} and forces {{beta}} to be
resolved to {{3}}. Both of these pairs then become {{()}} and be absorbed. We
are left with the pair {{(true,3)}} as desired.

>>>>

%% I am getting an error about too many math packages used!!!
%% I commented the ones we don't seem to be using.

\usepackage{graphicx}
%%\usepackage{longtable}
%%\usepackage{comment}
\usepackage{amsmath}
%%\usepackage{mdwlist}
%%\usepackage{txfonts}
\usepackage{xspace}
%%\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
%%\usepackage[nodayofweek]{datetime}
%%\usepackage{etex}
%%\usepackage[all, cmtip]{xy}
%%\usepackage{xcolor}
%%\usepackage{listings}
%%\usepackage{bm}
\usepackage{cmll}

@@

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^*
%! ^^^ = ^{-1}
%! eps* = \epsilon^*
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

Continuations made their introduction to the world of programming language
semantics as a mathematical device to model first-class labels and
jumps~\cite{springerlink:10.1023/A:1010026413531}. In a remarkable
development, Filinski~\cite{Filinski:1989:DCI:648332.755574} observed that
--- with the right perspective --- this highly imperative concept was
actually the symmetric dual of values. The heart of the observation is that
values represent entities that flow from producers to consumers while
continuations represent \emph{demands} for such entities, that flow from
consumers to producers. In more detail, Filinski describes continuations as
representing ``the \emph{lack} or \emph{absence} of a value, just as having a
negative amount of money represents a debt.'' He then proceeds to construct a
language where values and continuations are treated truly symmetrically. To
that end, he abandons the $\lambda$-calculus amalgamation of functions and
values and distinguishes between three different syntactic classes:
functions, values, and continuations, with the property that any function can
be used either as a value transformer or as a continuation transformer.

This highly intuitive and appealing idea was further explored and refined by
many authors~\cite{Griffin:1989:FNC:96709.96714, curien00duality,
  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet despite of its appeal, the idea
that continuations of this

appeared as functional
representation of labels and jumps,
Filinski, in 1989, observed that they
are the categorical introduced the symmetric $\lambda$-calculus

Computationally, symmetry exhibits itself as a duality between
two concepts.

Symmetry is pervasive in both natural and man-made environments.

In 1989, Filinski~\cite{Filinski:1989:DCI:648332.755574} observed that values
and continuations are dual notions. This observation was followed by numerous

Rauszer~\cite{springerlink:10.1007/BF02120864,rauszer,rauszer2} introduced a
logic which contains a dual to implication. Her work has been distilled in
the form of \emph{subtractive logic}~\cite{Crolard01} which has recently been
related to coroutines~\cite{Crolard01082004} and delimited
continuations~\cite{Ariola:2009:TFD:1743339.1743381}.

@@

We show a deep symmetry between functions and delimited continuations, values
and continuations that arises in {{langRev}} in a manner that is reminiscent
of Filinksi's Symmetric \lcal ~\cite{Filinski:1989:DCI:648332.755574}. The
symmetry arises by extending {{langRev}} with a notion of additive duality
over the monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
Compact Closed Categories. The resulting dual types, which we denote {{-b}},
have a time traveling ``backward information flow'' interpretation and allow
for the encoding of higher-order function and iteration via the construction
of trace operators, thereby making the extended language {{langRevEE}} a
Turing-complete reversible programming language with higher-order functions
and first-class delimited continuations.

@@

The goal is to produce a value of type $(\texttt{bool} \times
\texttt{int})$. One part of the program, $e_1$, knows how to produce the
value of type $\texttt{int}$ (say {{3}}) and another part, $e_2$, knows how
to produce the value of type $\texttt{bool}$ (say true). Computationally, we
model this as follows. The first subcomputation $e_1$ produces {{((),3)}} and
uses the isomorphism to convert {{()}} to $(\alpha,1/\alpha)$ where $\alpha$
is a yet-unknown boolean value. Reshuffling the components, $e_1$ produces
$((\alpha,3), 1/\alpha)$. Similarly, $e_2$ can independently produce
$((true,\beta),1/\beta)$ where $\beta$ is a yet-unknown integer value.  When
the two values meet, we can group the components as follows:
\[
(\alpha,\beta)  \qquad (3,1/\beta)  \qquad (true,1/\alpha)
\]
Using the isomorphism in the right-to-left direction on the last two pairs,
forces $\alpha$ to be resolved to true and forces $\beta$ to be resolved to
{{3}}. Both of these pairs then become {{()}} and be absorbed. We are left
with the pair $(true,3)$ as desired.
